---
title: "Phase1 Markdown"
author: "Duarte Da Silva"
date: "2025-04-27"
output: html_document
---
# 📈 Trade Signal Generator Using MACD + RSI

### A modular R function that scans for bullish/bearish entry setups in any stock

# Introduction

This project was built to simulate how a technical trader might use MACD and RSI conditions — in combination with price positioning — to evaluate entry signals. I coded each condition check manually using R, and wrapped the logic in modular functions for long-term usability.

This is a personal passion project I set out to do as a complimentary assignment to apply the knowledge learned on the Datacamp track Introduction to R for Finance, my objectives for today are the following:

-   Write down important theory notes
-   Pull Stock Data
-   Plot a Candlestick Chart
-   Calculate Indicators
-   Plot indicators
-   Risk Management Function
-   Wrap up Day 1.

```{r}
# Sunday, April 27th 2025 Completing Phase 1 of R for Finance project

if(!require("quantmod")) install.packages("quantmod")
if(!require("tidyverse")) install.packages("tidyverse")
if(!require("ggplot2")) install.packages("ggplot2")
if(!require("TTR")) install.packages("TTR")
```

```{r}
library(quantmod)
library(tidyverse)
library(TTR)
library(ggplot2)
```

# Day 1 Sunday, April 27th

## Market Theory Notes (for Phase 1)

### 1. Supply and Demand

####Technical Definition:

Markets move because of changes in supply (how much of something is available) and demand (how much people want it).

If demand > supply, prices rise (buyers fight for limited supply).

If supply > demand, prices fall (too much of something, buyers become picky).

####Simple Example:

Imagine there are only 3 ice cream cones at the playground, but 10 kids want one.
Kids will start offering more and more to get one — price goes up! 

But if there are 100 ice cream cones and only 3 kids, the cones sit around — prices drop because nobody is desperate.

### 2. Catalysts

####Technical Definition:

A catalyst is an event that suddenly increases supply or demand.

Examples:

Earnings Reports: Company profits higher than expected → more buyers.

News Events: New product launch, lawsuit, scandal, CEO resigns.

Economic Reports: Inflation data, employment rates, interest rate hikes.

####Simple Example:

Imagine if you’re selling lemonade.
Suddenly it's a hot day and everyone’s thirsty — BOOM, everyone wants your lemonade = catalyst.
Or if it starts raining and people run home — nobody buys lemonade = bad catalyst.

### 3. Types of Orders

####Technical Definition:

Traders use different kinds of orders to control how and when they buy or sell:

Market Order: Buy/sell immediately at current price.

Limit Order: Buy/sell only if price reaches a specific number.

Stop Order: Trigger a buy/sell when price passes a threshold.

####Simple Example:

Market Order = "I want a toy NOW, I’ll pay whatever the price is." 

Limit Order = "I’ll only buy the toy if it's $10 or less." 

Stop Order = "If the toy’s price falls to $5, tell me and I’ll buy it." 

### 4. Risk Concepts

####Technical Definition:

Position Sizing: Deciding how much of your money you put into a single trade.

Stop Loss: A pre-set rule that closes your trade if you lose too much money.

Risk per Trade: How much (%) of your total account you are willing to risk on one bet.

Key rules:

Good traders risk only 1–2% of their money per trade.

Stop Losses protect you from devastating losses.

####Simple Example:

Imagine you have 100 dollars.

You say, “I’ll only bet 2 dollars on this game of rock-paper-scissors.”

If you lose, you lose 2 dollars — no big deal.

If you lose 10 games, you still have 80 dollars left.

Without rules, you could lose all your money in one bad bet!

## 1. Pulling Stock Data & Plotting Charts

```{r, eval = FALSE, include=FALSE}
# Get data on APPL stock using getSymbols() from Yahoo Finance.

getSymbols("AAPL") 

chartSeries(AAPL,theme = chartTheme("white"))

```

```{r}
# You can search a specific timeframe with the following syntax: 

getSymbols("AAPL", src = "yahoo", from = "2023-01-01")

# Plot candlestick chart 
chartSeries(AAPL, theme = chartTheme("white"))

# Add Indicators

# 1. RSI
addRSI(n = 14) # 14 is the standard period for RSI

# 2. SMA20 & SMA50
addSMA(n = 20)
addSMA(n = 50)

# 3. MACD 
addMACD(fast = 12, slow =26, signal = 9, type = "EMA")

# 4. EMA (Optional) 

addEMA(n = 20) 
```

## Creating a function()

The function() is a machine or function that takes inputs, performs an operation with them based on instructions and gives you an output, for example: 

```{r, eval = FALSE}
function_name <- function(inputs) {
  # Code to process the inputs 
  # Output (usually with return())
}
```

* function_name --  the name for your machine.

* <- function(inputs) -- Tell R you're creating a function and list what information it needs.

* {} Write instructions for your machine.

#### Simple Example 

```{r}
say_hello <- function(name) { 
  message <- paste("Hello", name)
  return(message)
}

say_hello("Oscar")
```

* Name is the input. 
* The function builds a message using paste().
* return() sends the message back to you.
* paste() combines inputs into a text string.
* Using return() is necessary for R to print the result.

## 2. Calculating position size

```{r}
 # 1. Calculate position size
calculate_position_size <- function(account_balance, risk_percent, stop_loss_percent) { 
  
  # 2. Calculate $ amount to risk per trade 
  dollar_risk <- account_balance * (risk_percent / 100)
  
  # 3.  Calculate position size 
  position_size <- dollar_risk / (stop_loss_percent / 100)

  return(list(
    position_size = position_size,
    dollar_risk = dollar_risk
  ))
  }
```

-   account_balance = total money you have in your account (e.g., \$10,000).

-   risk_percent = how much % of your account you are willing to lose in one trade (e.g., 1%).

- dollar_risk = the exact amount of real $ you are willing to lose. i. e: 500 * (1/100) so 500 * 0.01 = 5.

- position_size = The amount of stock youre buying in the operation, i. e: 5 / (2/100) so 5 / 0.02 = 250, meaning you can fit the value "0.02" 250 times in the number 5.

-   stop_loss_percent = how far away your stop loss is from entry price (e.g., 2%).

#### Test Cases

```{r}

# Test Case 1 
calculate_position_size(500, 1, 2)

# Test Case 2 
calculate_position_size(50, 2, 5)

# Test Case 3 
calculate_position_size(5000, 1, 1.5)

# Test Case 4 
calculate_position_size(25000, 0.5, 3)
```

## Wrapping up Day 1 

Today I learned to:

* Use getSymbols() function to import data from Yahoo finance.

* Use chartSeries() and addSMA(), etc to plot charts and add indicators such as: SMA, EMA, MACD and RSI.

* Build a function to calculate the dollar risk and position size of a possible trade.

# Day 2 Monday, April 28th

## 3. Building Trade Signal Detection Function

The objective for day two is to build a function that: 

* Suggests possible trade setups.
* Detects if stock is near Support/Resistance levels based on SMA.
* Confirms entry signal guided by a trigger combo of RSI and MACD.
* Calculates position size & possible $ loss.


To build this function im going to complete the next steps: 

 1. Pull the stock data
 2. Calculate Indicators
 3. Create a Trade Setup Logic
 4. Confirm with RSI and MACD
 5. Create Output
 
```{r Day 2 work}

RSI_MACD_combo <- function(ticker) {
  library(quantmod)
# 1. Pull Stock Data 

stock_data <- getSymbols(ticker, from = "2024-04-28", auto.assign = TRUE)
stock_data <- get(ticker)

# 2. Calculate indicators and save them as objects

sma20 <- SMA(Cl(stock_data), n = 20) 
sma50 <- SMA(Cl(stock_data), n = 50)
rsi14 <- RSI(Cl(stock_data), n = 14) 
macd_vals <- MACD(Cl(stock_data))

# 3.  Build condition checks 

close_today <- as.numeric(last(Cl(stock_data)))

# 4. Today's SMA values
sma20_today <- as.numeric(last(sma20))
sma50_today <- as.numeric(last(sma50))

## Absolute difference between closing price and SMA

diff_sma20 <- abs(close_today - sma20_today)
diff_sma50 <- abs(close_today - sma50_today)

## Calculate percentage difference relative to SMA

perc_diff_sma20 <- diff_sma20 / sma20_today
perc_diff_sma50 <- diff_sma50 / sma50_today

## Check if the percentage difference is less than 2% 

near_sma20 <- perc_diff_sma20 <= 0.02
near_sma50 <- perc_diff_sma50 <= 0.02

## Combined condition: near ANY SMA 

near_any_sma <- near_sma20 | near_sma50

## Print both SMA condition results

print(near_sma20)
print(near_sma50)

# 5. Today's RSI value

rsi_today <- as.numeric(last(rsi14))

## Check RSI conditions

rsi_buy_signal <- rsi_today < 30
rsi_sell_signal <- rsi_today > 70

## Print results
print(rsi_today)
print(rsi_buy_signal)
print(rsi_sell_signal)

# 6. MACD cross detection

# WHAT YOU WANT TO CHECK: 
## Buy signal: when MACD crosses above signal line.
## Sell signal: when MACD crosses below signal line.

## Extract MACD line and Signal line separately

macd_line <- macd_vals[, 1]
macd_signal <- macd_vals[, 2]

## Get today's and yesterday's values

macd_today <- as.numeric(last(macd_line))
signal_today <- as.numeric(last(macd_signal))

macd_yesterday <- as.numeric(last(macd_line[nrow(macd_line) - 1]))
signal_yesterday <- as.numeric(macd_signal[nrow(macd_signal) - 1])

## Check if a bullish crossover happened
macd_bullish_cross <- (macd_yesterday < signal_yesterday) & (macd_today > signal_today)

## Check if a bearish crossover happened
macd_bearish_cross <- (macd_yesterday > signal_yesterday) & (macd_today < signal_today)

## Print results 
print(macd_bullish_cross)
print(macd_bearish_cross)

# 7. Build Trade Desicion Logic

if(near_any_sma & rsi_buy_signal & macd_bullish_cross) { 
  trade_signal <- "LOOK INTO"
} else if (near_any_sma & rsi_sell_signal & macd_bearish_cross) { 
  trade_signal <- "LOOK INTO"
} else { 
    trade_signal <- "NO ACTION"
}


 
  sma_20_signal <- near_sma20
  sma_50_signal <- near_sma50
  any_sma_signal  <- near_any_sma

 
  # Final Output
  
  print(paste("Ticker:", ticker))
  print(paste("Trade Recommendation:", trade_signal))
  print(paste("Price is near SMA 20:", near_sma20))
  print(paste("Price is near SMA 50:", near_sma50))
  print(paste("Today's Close:", round(close_today,2)))
  print(paste("RSI Today:", round(rsi_today)))
  
return(trade_signal)
}


```

```{r}
RSI_MACD_combo("META")
```

## Wrapping Up Day 2 

<details>
<summary><strong>Click to expand: Trade Signal Explanation</strong></summary>

 Trade Signal Function Progress

Today I spent most of my time working on the Trade Signal Function, which I named RSI_MACD_combo. I'm documenting this process to review my work and make sure I understand every part of it as I continue to build and evolve the function.

The goal of this function is simple:
➡️ Use RSI and MACD triggers to suggest potential trade signals.

  Step 1: Pulling the Data
I used getSymbols() to pull data directly from Yahoo Finance.
To make the function dynamic and reusable, I used a ticker input instead of hard-coding any stock names. The pulled data is stored as stock_data:



stock_data <- getSymbols(ticker, from = "2024-04-28", auto.assign = FALSE)

  Step 2: Adding Indicators
I calculated all indicators based on the stock’s closing prices, using:

SMA() for the 20- and 50-day moving averages,

RSI() for momentum signals (14-day),

MACD() for crossover detection.

These all used the Cl() function to extract closing prices from the stock data.

 Step 3: Getting Today’s Values
To check current market conditions, I used last() and as.numeric() to store today’s key values:



close_today <- as.numeric(last(Cl(stock_data)))
sma20_today <- as.numeric(last(sma20))
sma50_today <- as.numeric(last(sma50))

Step 4: Distance from SMA Check
I used abs() to calculate the absolute difference between the closing price and each SMA. Then I calculated the percentage difference to define whether the price is "near" either moving average. I considered within 2% to be "close."

 Step 5: RSI Signal
I grabbed today's RSI value using:



rsi_today <- as.numeric(last(rsi14))
Then I defined:

RSI below 30 = potential buy signal

RSI above 70 = potential sell signal

 Step 6: MACD Crossover
I separated the MACD line and Signal line from the MACD output, then built two condition checks:

Bullish Cross = MACD crosses above signal line

Bearish Cross = MACD crosses below signal line

This helps identify momentum shifts in price action.

 Step 7: Trade Decision Logic
Once I had all the indicators and condition checks in place, I built the actual decision logic.

The function checks:

Is the price near either SMA?

Is there a buy/sell signal from RSI?

Is there a MACD crossover confirming that signal?

If all conditions align, the function returns "LOOK INTO", along with:

The ticker,

The RSI reading,

The closing price,

Which SMA the price is closest to.

  This project is helping me improve both my R syntax and my ability to think like a data-driven trader.
Every time I rebuild a piece of the function, I understand it a little better — and I’m excited to keep refining it over the next few days.

</details>

